Why is Stochastic Lambda > Deterministic?

### 2022-04-26 notes/understanding stochastic lambda.Rmd
update notes on stochastic lambda

@@

withr::with_dir(here(),{
  tar_load(data_full)
  tar_load(vit_other_ff)
})
```

@@

```{r warning=FALSE, class.source = 'fold-hide'}
single_years <-   
  data_full %>%
  filter(year>1999) %>% 
  select(-L, -spei_history) %>% 
  group_by(year_fct) %>% 
  group_split()

vit_list_yr <- 
  single_years %>% 
  filter(habitat == "1-ha") %>% 
  map(~{
    vit_list_det_ff = c(
      list(
        vit_surv = surv_det(.x),
        vit_size = size_det(.x),
        vit_flwr = flwr_det(.x),
        vit_size_sdlg = size_sdlg_det(.x),
        vit_surv_sdlg = surv_sdlg_det(.x)
      ),
      vit_other_ff)
    })

ipms_single_yr <-
  map(vit_list_yr, ~{
    make_proto_ipm_det(.x) %>% 
      make_ipm(iterations = 1000,  #only needs 100 to converge
               normalize_pop_size = FALSE, # to run as PVA
               usr_funs = list(get_scat_params = get_scat_params))
  })

lambdas <- ipms_single_yr %>% map_dbl(lambda)
```

```{r warning=FALSE}
#worst year
(worst_year <- c(2000:max(data_full$year))[which.min(lambdas)]); min(lambdas)

#best year
(best_year <- c(2000:max(data_full$year))[which.max(lambdas)]); max(lambdas)

two_yrs <- data_full %>% filter(year %in% c(best_year, worst_year))
```

@@

```{r warning=FALSE, class.source = 'fold-hide'}
vit_list_det_ff = c(
  list(
    vit_surv = surv_det(two_yrs),
    vit_size = size_det(two_yrs),
    vit_flwr = flwr_det(two_yrs),
    vit_size_sdlg = size_sdlg_det(two_yrs),
    vit_surv_sdlg = surv_sdlg_det(two_yrs)
  ),
  vit_other_ff)

ipm_det_ff = make_proto_ipm_det(vit_list_det_ff) %>% 
  make_ipm(iterations = 1000,  #only needs 100 to converge
           normalize_pop_size = FALSE, # to run as PVA
           usr_funs = list(get_scat_params = get_scat_params))
lambda(ipm_det_ff)
```

@@

```{r warning=FALSE, class.source = 'fold-hide'}
## forets fragments
vit_list_stoch_ff = c(
  list(
    vit_surv = surv_raneff(two_yrs),
    vit_size = size_raneff(two_yrs),
    vit_flwr = flwr_raneff(two_yrs),
    vit_size_sdlg = size_sdlg_raneff(two_yrs),
    vit_surv_sdlg = surv_sdlg_raneff(two_yrs)
  ), 
  vit_other_ff)

@@

  # define implementation with midpoint rule
  define_impl(make_impl_args_list(
    kernel_names = c("P_yr",        "go_sdlg_yr", "stay_sdlg", "leave_sdlg_yr"),
    int_rule = "midpoint",
    state_start  = c("log_size", "log_size", "sdlg",     "sdlg"),
    state_end    = c("log_size", "sdlg",     "sdlg",     "log_size")
  )) %>% 
  #define lower bound, upper bound, and number of meshpoints for log_size
  define_domains(log_size = c(0, 8.018296, 100)) %>%  
  #arbitrary starting population state
  define_pop_state(n_log_size = runif(100), n_sdlg = .1)
```

@@

```{r warning=FALSE}
ipm_stoch_ff <- 
  proto_ipm_stoch_ff %>%
  make_ipm(iterations = 1000,
           normalize_pop_size = FALSE, # to run as PVA
           usr_funs = list(get_scat_params = get_scat_params))
```

@@

```{r}
lambda(ipm_det_ff) 
lambda(ipm_stoch_ff, log = FALSE, burn_in = 0.2)
```

Yes, stochastic lambda is still a bit higher!

What about the averages for each year?
Are they consistent with a bad and good year?

```{r}
all_lams <- 
  tibble(year = ipm_stoch_ff$env_seq,
         lambda = lambda(ipm_stoch_ff, type_lambda = "all")[,1])
all_lams %>%
  group_by(year) %>%
  summarize(mean(lambda))
```

Yes, but both the good and bad years mean lambdas are *greater* than the deterministic lambda.

How do the mean lambdas from the good and bad year compare to the deterministic lambdas from individual years?

```{r}
max(lambdas)
min(lambdas)
```

@@

#deterministic, just 2001
det_fitted_2001 <- 
  vit_list_yr[[which.max(lambdas)]] %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>% 
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. 2001")

#deterministic, just 2003
det_fitted_2003 <- 
  vit_list_yr[[which.min(lambdas)]] %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>% 
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. 2003")

#deterministic with both years combined
det_fitted <- 
  vit_list_det_ff %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. both years")

#stochastic
stoch_fitted <- 
  vit_list_stoch_ff %>%
  map_df(~safe_predict(.x, newdata= newdata, type = "response")) %>%
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = paste("stoch.", year_fct))

bind_rows(stoch_fitted, det_fitted, det_fitted_2001, det_fitted_2003) %>% 
  ggplot(aes(x = log_size_prev, y = .fitted, color = year_fct, linetype = year_fct)) +
  geom_line(alpha = 0.9) +
  facet_wrap(~vital_rate, scales = "free")

>>>>

# TODO

-   Get code to run

-   Instead of deterministic IPM using different vital rates models, use the random effects models, but plug in `newdata = (log_size_prev = log_size, year = "newlevel")` to get mean fitted values.

-   add analysis with CF

-   Don't normalize population sizes and plot both CF and FF as PVA for both models

-   Try year as fixed effect

-   Read Rees & Ellner 2009.
    Consider either fixed effect of year + kernel sampling OR random effect of year + parameter sampling.
    Maybe no added value for random effect of year + kernel sampling??

Rees, Mark, and Stephen P. Ellner.
"Integral Projection Models for Populations in Temporally Varying Environments." Ecological Monographs 79, no. 4 (2009): 575--94.
<https://doi.org/10.1890/08-1474.1>.

@@

withr::with_dir(here(),{
  tar_load(c(
    data_full,
    pop_vec_ff,
    pop_vec_cf,
    vit_other_ff,
    vit_other_cf
  ))
})

#don't need matrix columns for this
data_full <- data_full %>% select(-L, -spei_history)
```

@@

```{r warning=FALSE, class.source = 'fold-hide'}
single_years <-   
  data_full %>%
  filter(year>1999) %>% 
  group_by(year_fct) %>% 
  group_split()

#fit vital rates for FF  
vit_list_yr_ff <- 
  single_years %>% 
  map(~dplyr::filter(.x, habitat == "1-ha")) %>% 
  map(~{
    vit_list_det = c(
      list(
        vit_surv = surv_det(.x),
        vit_size = size_det(.x),
        vit_flwr = flwr_det(.x),
        vit_size_sdlg = size_sdlg_det(.x),
        vit_surv_sdlg = surv_sdlg_det(.x)
      ),
      vit_other_ff)
    })
# make IPMs for FF
ipms_yr_ff <-
  map(vit_list_yr_ff, ~{
    make_proto_ipm_det(.x, pop_vec = pop_vec_ff) %>% 
      make_ipm(iterations = 1000,  #only needs 100 to converge
               normalize_pop_size = FALSE, # to run as PVA
               usr_funs = list(get_scat_params = get_scat_params))
  })

# fit vital rates for CF
vit_list_yr_cf <- 
  single_years %>% 
  map(~dplyr::filter(.x, habitat == "CF")) %>% 
  map(~{
    vit_list_det = c(
      list(
        vit_surv = surv_det(.x),
        vit_size = size_det(.x),
        vit_flwr = flwr_det(.x),
        vit_size_sdlg = size_sdlg_det(.x),
        vit_surv_sdlg = surv_sdlg_det(.x)
      ),
      vit_other_cf)
    })

#make IPMs for CF
ipms_yr_cf <-
  map(vit_list_yr_cf, ~{
    make_proto_ipm_det(.x, pop_vec = pop_vec_cf) %>% 
      make_ipm(iterations = 1000,  #only needs 100 to converge
               normalize_pop_size = FALSE, # to run as PVA
               usr_funs = list(get_scat_params = get_scat_params))
  })

#lambdas
lambdas_ff <- ipms_yr_ff %>% map_dbl(lambda)
lambdas_cf <- ipms_yr_cf %>% map_dbl(lambda)
```

```{r warning=FALSE}
#worst year
(worst_year_ff <- c(2000:max(data_full$year))[which.min(lambdas_ff)]); min(lambdas_ff)
(worst_year_cf <- c(2000:max(data_full$year))[which.min(lambdas_cf)]); min(lambdas_cf)

#best year
(best_year_ff <- c(2000:max(data_full$year))[which.max(lambdas_ff)]); max(lambdas_ff)
(best_year_cf <- c(2000:max(data_full$year))[which.max(lambdas_cf)]); max(lambdas_cf)

plot(2000:max(data_full$year), lambdas_cf, type = "l", col = "red"); lines(x =2000:max(data_full$year), y = lambdas_ff, col = "blue", type = "l")
```

2003 was the worst year in both habitats.
2001 was a good year in both habitats (but not the best in CF)

```{r warning=FALSE}
#combine good and bad year
two_yrs <- data_full %>% filter(year %in% c(2003, 2001))
two_yrs_ff <- two_yrs %>% filter(habitat == "1-ha")
two_yrs_cf <- two_yrs %>% filter(habitat == "CF")
```

@@

Fragments:

```{r warning=FALSE, class.source = 'fold-hide'}
vit_list_det_ff = c(
  list(
    vit_surv = surv_det(two_yrs_ff),
    vit_size = size_det(two_yrs_ff),
    vit_flwr = flwr_det(two_yrs_ff),
    vit_size_sdlg = size_sdlg_det(two_yrs_ff),
    vit_surv_sdlg = surv_sdlg_det(two_yrs_ff)
  ),
  vit_other_ff)

ipm_det_ff = make_proto_ipm_det(vit_list_det_ff, pop_vec = pop_vec_ff) %>% 
  make_ipm(iterations = 1000,  #only needs 100 to converge
           normalize_pop_size = FALSE, # to run as PVA
           usr_funs = list(get_scat_params = get_scat_params))
# FF
lambda(ipm_det_ff)
```

CF:

```{r warning=FALSE, class.source = 'fold-hide'}
vit_list_det_cf = c(
  list(
    vit_surv = surv_det(two_yrs_cf),
    vit_size = size_det(two_yrs_cf),
    vit_flwr = flwr_det(two_yrs_cf),
    vit_size_sdlg = size_sdlg_det(two_yrs_cf),
    vit_surv_sdlg = surv_sdlg_det(two_yrs_cf)
  ),
  vit_other_cf)

ipm_det_cf = make_proto_ipm_det(vit_list_det_cf, pop_vec = pop_vec_cf) %>% 
  make_ipm(iterations = 1000,  #only needs 100 to converge
           normalize_pop_size = FALSE, # to run as PVA
           usr_funs = list(get_scat_params = get_scat_params))
lambda(ipm_det_cf)
```

@@

```{r warning=FALSE, class.source = 'fold-hide'}
## forests fragments
vit_list_stoch_ff = c(
  list(
    vit_surv = surv_raneff(two_yrs_ff),
    vit_size = size_raneff(two_yrs_ff),
    vit_flwr = flwr_raneff(two_yrs_ff),
    vit_size_sdlg = size_sdlg_raneff(two_yrs_ff),
    vit_surv_sdlg = surv_sdlg_raneff(two_yrs_ff)
  ), 
  vit_other_ff)

@@

  # define implementation with midpoint rule
  define_impl(make_impl_args_list(
    kernel_names = c("P_yr",        "go_sdlg_yr", "stay_sdlg", "leave_sdlg_yr"),
    int_rule = "midpoint",
    state_start  = c("log_size", "log_size", "sdlg",     "sdlg"),
    state_end    = c("log_size", "sdlg",     "sdlg",     "log_size")
  )) %>% 
  #define lower bound, upper bound, and number of meshpoints for log_size
  define_domains(log_size = c(0, 8.018296, 100)) %>%  
  #arbitrary starting population state
  define_pop_state(pop_vectors = pop_vec_ff)
```

```{r warning=FALSE, class.source = 'fold-hide'}
## forests fragments
vit_list_stoch_cf = c(
  list(
    vit_surv = surv_raneff(two_yrs_cf),
    vit_size = size_raneff(two_yrs_cf),
    vit_flwr = flwr_raneff(two_yrs_cf),
    vit_size_sdlg = size_sdlg_raneff(two_yrs_cf),
    vit_surv_sdlg = surv_sdlg_raneff(two_yrs_cf)
  ), 
  vit_other_cf)

proto_ipm_stoch_cf <- 
  init_ipm(
    sim_gen = "general",
    di_dd = "di",
    det_stoch = "stoch",
    kern_param = "kern"
  ) %>%

    data_list = vit_list_stoch_cf,
    states          = list(c('log_size')),
    uses_par_sets   = TRUE,
    par_set_indices = list(yr = c(2001, 2003)),
    evict_cor       = TRUE,
    evict_fun       = truncated_distributions("t.scaled", "G_z1z")
  ) %>%

    data_list     = vit_list_stoch_cf,
    states        = list(c('log_size', 'sdlg')),
    uses_par_sets = TRUE,
    par_set_indices = list(yr = c(2001, 2003))
  ) %>%

    data_list = vit_list_stoch_cf,
    states = list(c("sdlg", "log_size")),
    uses_par_sets = TRUE,
    par_set_indices = list(yr = c(2001, 2003)),
    evict_cor = TRUE,
    evict_fun = truncated_distributions("t.scaled", "G_z2_sdlg")
  ) %>%

  # define implementation with midpoint rule
  define_impl(make_impl_args_list(
    kernel_names = c("P_yr",        "go_sdlg_yr", "stay_sdlg", "leave_sdlg_yr"),
    int_rule = "midpoint",
    state_start  = c("log_size", "log_size", "sdlg",     "sdlg"),
    state_end    = c("log_size", "sdlg",     "sdlg",     "log_size")
  )) %>% 
  #define lower bound, upper bound, and number of meshpoints for log_size
  define_domains(log_size = c(0, 8.018296, 100)) %>%  
  #arbitrary starting population state
  define_pop_state(pop_vectors = pop_vec_cf)
```

@@

```{r warning=FALSE}
ipm_stoch_ff <- 
  proto_ipm_stoch_ff %>%
  make_ipm(iterations = 1000,
           normalize_pop_size = FALSE, # to run as PVA
           usr_funs = list(get_scat_params = get_scat_params))

ipm_stoch_cf <- 
  proto_ipm_stoch_cf %>%
  make_ipm(iterations = 1000,
           normalize_pop_size = FALSE, # to run as PVA
           usr_funs = list(get_scat_params = get_scat_params))
```

@@

```{r}
all_lams <-
  list(ff = ipm_stoch_ff, cf = ipm_stoch_cf) %>% 
  map_df(
    ~tibble(year = .x$env_seq,
         lambda = lambda(.x, type_lambda = "all")[,1])%>% 
  #remove burnin
  slice(101:n()),
  .id = "habitat"
  )

```{r lambdas}
tribble(~model, ~habitat, ~year, ~lambda,
        "det", "ff", "both combined", lambda(ipm_det_ff),
        "stoch", "ff", "both combined", lambda(ipm_stoch_ff, log = FALSE),
        "det", "cf", "both combined", lambda(ipm_det_cf),
        "stoch", "cf", "both combined", lambda(ipm_stoch_ff, log = FALSE),
        "det", "ff", "2001", lambdas_ff[2],
        "det", "ff", "2003", lambdas_ff[4],
        "det", "cf", "2001", lambdas_cf[2],
        "det", "cf", "2003", lambdas_cf[4]
) %>% 
  bind_rows(
    all_lams %>%
      group_by(habitat, year) %>%
      summarize(lambda = mean(lambda)) %>% 
      mutate(year = paste(year, "mean"),
             model = "stoch")
  ) %>% arrange(model, habitat)
```

Stochastic lambda is still slightly higher than deterministic

Within stochastic IPMs, the mean lambdas for individual transition years are consistent with good and bad years, but both are *higher* than deterministic lambda with both years combined!

@@

#deterministic, just 2001
det_fitted_2001_ff <- 
  vit_list_yr_ff[[2]] %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>% 
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. 2001")

#deterministic, just 2003
det_fitted_2003_ff <- 
  vit_list_yr_ff[[4]] %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>% 
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. 2003")

#deterministic with both years combined
det_fitted_ff <- 
  vit_list_det_ff %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. both years")

#stochastic
stoch_fitted_ff <- 
  vit_list_stoch_ff %>%
  map_df(~safe_predict(.x, newdata= newdata, type = "response")) %>%
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = paste("stoch.", year_fct))

bind_rows(stoch_fitted_ff, det_fitted_ff, det_fitted_2001_ff, det_fitted_2003_ff) %>% 
  ggplot(aes(x = log_size_prev, y = .fitted, color = year_fct, linetype = year_fct)) +
  geom_line(alpha = 0.9) +
  facet_wrap(~vital_rate, scales = "free") +
  labs(title = "Forest Fragments")

#deterministic, just 2001
det_fitted_2001_cf <- 
  vit_list_yr_cf[[2]] %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>% 
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. 2001")

#deterministic, just 2003
det_fitted_2003_cf <- 
  vit_list_yr_cf[[4]] %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>% 
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. 2003")

#deterministic with both years combined
det_fitted_cf <- 
  vit_list_det_cf %>% 
  map_df(~safe_predict(.x, newdata = newdata, type = "response")) %>% 
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = "det. both years")

#stochastic
stoch_fitted_cf <- 
  vit_list_stoch_cf %>%
  map_df(~safe_predict(.x, newdata= newdata, type = "response")) %>%
  bind_cols(newdata) %>%
  select(-vit_fruits, -vit_seeds, -vit_germ_est) %>% 
  pivot_longer(starts_with("vit_"), names_to = "vital_rate", values_to = ".fitted", names_prefix = "vit_") %>% 
  mutate(year_fct = paste("stoch.", year_fct))

bind_rows(stoch_fitted_cf, det_fitted_cf, det_fitted_2001_cf, det_fitted_2003_cf) %>% 
  ggplot(aes(x = log_size_prev, y = .fitted, color = year_fct, linetype = year_fct)) +
  geom_line(alpha = 0.9) +
  facet_wrap(~vital_rate, scales = "free") +
  labs(title = "Continuous Forest")
